---
title: "GLOSSA regional"
subtitle: "Version 0.1.0"
author: "Jorge Mestre Tomás"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{GLOSSA regional}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# Load dependencies

```{r}
# Load libraries
library(glossa)
library(dplyr)
library(ggplot2)
```

# Load oceania polygon

```{r}
sf::sf_use_s2(FALSE)
study_area_poly <- rnaturalearth::ne_countries(continent = "Oceania", scale = "medium", returnclass = "sf") %>% 
  sf::st_as_sfc() %>%
  sf::st_union() %>%
  sf::st_make_valid() %>%
  sf::st_wrap_dateline() %>% 
  sf::st_crop(c(xmin = 112.90820, ymin =-54.74922, xmax = 180.00000, ymax = 18.80679)) %>% 
  glossa::invert_polygon()

non_study_area_poly <- glossa::invert_polygon(study_area_poly)
```

# Load presence(/absence) data and environmental layers

```{r}
# Load presence(/absence) data
pa_file <- "../inst/extdata/sp1.csv"
raw_pa <- glossa::read_glossa_pa(pa_file)

# Load historical layers
historical_files <- "../inst/extdata/historical_layers.zip"
past_layers <- glossa::read_glossa_layers(historical_files)

# Load future layers
future_files <- "../inst/extdata/future1_layers.zip"
future_layers <- glossa::read_glossa_layers(future_files)
```

```{r, echo = FALSE}
ggplot() +
  geom_sf(data = non_study_area_poly, color = "#353839", fill = "antiquewhite") +
  geom_point(data = raw_pa, aes(x = decimalLongitude, y = decimalLatitude, color = "#F6733A")) +
  theme(
    panel.grid.major = element_line(
      color = gray(.5),
      linetype = "dashed",
      linewidth = 0.5
    ),
    panel.background = element_rect(fill = "white"),
    axis.title = element_blank(),
    legend.position = "bottom"
  )
```

# Clean coordinates

```{r}
coords <- c("decimalLongitude", "decimalLatitude")

# Remove NA coordinates
clean_pa <- raw_pa[complete.cases(raw_pa[, coords]), ]

# Round coordinates to 4 decimal digits (this is up to your choice and the precision you need)
decimal_digits <- 4
clean_pa[, coords[1]] <- round(clean_pa[, coords[1]], decimal_digits)
clean_pa[, coords[2]] <- round(clean_pa[, coords[2]], decimal_digits)

# Remove duplicated points
clean_pa <- glossa::remove_duplicate_points(clean_pa, coords = coords)

# Remove points outside the ocean boundaries (overlapping land)
clean_pa <- glossa::remove_points_poly(clean_pa,
                                       sf_poly = study_area_poly,
                                       overlapping = FALSE,
                                       coords = coords)
```

You can summarize all these steps by just running the `glossa::clean_coordinates()` function:

```{r}
clean_pa <- glossa::clean_coordinates(
  data = raw_pa,
  sf_poly = study_area_poly,
  overlapping = FALSE,
  decimal_digits = decimal_digits,
  coords = coords
)
```

# Covariate layer processing

```{r}
past_layers <- lapply(past_layers, function(x){
  glossa::layer_mask(layers = x, sf_poly = study_area_poly)
})

future_layers <- lapply(future_layers, function(x){
  glossa::layer_mask(layers = x, sf_poly = study_area_poly)
})
```

```{r}
# Compute mean and sd of the historical time series for each environmental variable
past_mean <- lapply(past_layers, function(x){
  mean(as.vector(x), na.rm = TRUE)
})

past_sd <- lapply(past_layers, function(x){
  sd(as.vector(x), na.rm = TRUE)
})

# Scale layers with past mean and sd
past_layers <- lapply(seq_along(past_layers), function(x, n, i){
  terra::scale(x[[n[i]]], center = past_mean[[n[i]]], scale = past_sd[[n[i]]])},
  x = past_layers,
  n = names(past_layers))
names(past_layers) <- names(past_mean)

future_layers <- lapply(seq_along(future_layers), function(x, n, i){
  terra::scale(x[[n[i]]], center = past_mean[[n[i]]], scale = past_sd[[n[i]]])},
  x = future_layers,
  n = names(future_layers))
names(future_layers) <- names(past_mean)
```

```{r}
# Get mean of scaled past layers for model fitting
historical_layers <- lapply(past_layers, function(x){
  terra::mean(x, na.rm = TRUE)
})
historical_layers <- terra::rast(historical_layers)
```

# Build model matrix and pseudoabsences

```{r}
y_resp <- cbind(clean_pa,
                  terra::extract(historical_layers, clean_pa[, coords])) %>%
  tidyr::drop_na()  %>%
  dplyr::select(colnames(clean_pa))
```


```{r}
# Generate balanced random pseudoabsences
if (all(y_resp[, "pa"] == 1)){
  set.seed(1234)
  y_resp <- glossa::generate_pseudo_absences(y_resp, study_area_poly, historical_layers)
}
```


```{r}
ggplot2::ggplot() +
  tidyterra::geom_spatraster(data = historical_layers$x1) +
  ggplot2::scale_fill_gradientn(colours = c("#9fe5d7", "#65c4d8", "#39a6d5", "#2b8fc7", "#f67d33", "#f44934", "#ca3a43", "#9e0142")) +
  geom_sf(data = non_study_area_poly, color = "#353839", fill = "antiquewhite") +
  geom_point(data = y_resp, aes(x = decimalLongitude, y = decimalLatitude, color = as.factor(pa)), alpha = 1) +
  scale_color_manual(values = c("0" = "black","1" = "green"), labels = c("Absences", "Presences"), name = NULL) +
  theme(
    panel.grid.major = element_line(
      color = gray(.5),
      linetype = "dashed",
      linewidth = 0.5
    ),
    panel.background = element_rect(fill = "white"),
    axis.title = element_blank(),
    legend.position = "bottom"
  )
```

# Fit native range and suitable habitat model

```{r}
coords_layer <- glossa::create_coords_layer(historical_layers, study_area_poly, scale_layers = TRUE)
```

Now we can fit the model for the native range and the suitable habitat

```{r}
model_native_range <- glossa::fit_bart_model(
  y_resp,
  c(historical_layers, coords_layer),
  seed = 1234
)

model_suitable_habitat <- glossa::fit_bart_model(
  y_resp,
  historical_layers,
  seed = 1234
)
```

# Predict past, historical and future

```{r}
historical_native_range <- glossa::predict_bart(model_native_range, c(historical_layers, coords_layer))
historical_suitable_habitat <- glossa::predict_bart(model_suitable_habitat, historical_layers)
```

```{r}
# Past prediction
past_suitable_habitat <- lapply(1:terra::nlyr(past_layers[[1]]), function(i){
  # Stack covariates by year
  pred_layers <- lapply(past_layers, function(y){
    return(y[[i]])
  })
  pred_layers <- terra::rast(pred_layers)
  
  glossa::predict_bart(model_suitable_habitat, pred_layers)
})

future_suitable_habitat <- lapply(1:terra::nlyr(future_layers[[1]]), function(i){
  # Stack covariates by year
  pred_layers <- lapply(future_layers, function(y){
    return(y[[i]])
  })
  pred_layers <- terra::rast(pred_layers)
  
  glossa::predict_bart(model_suitable_habitat, pred_layers)
})
```



