---
title: "Step by step guide to GLOSSA analysis fitting BART models"
subtitle: "Version 0.1.0"
author: "Jorge Mestre Tomás"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Step by step guide to GLOSSA analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load libraries, functions and supp. data

```{r}
# * Load packages ----
list_of_packages <- c("tidyverse","terra", "tidyterra", "glossa")
req <- unlist(lapply(list_of_packages,require,character.only=TRUE))
need <- list_of_packages[req==FALSE]
if(length(need) > 0) {
  install.packages(need)
  lapply(need,require,character.only=TRUE)
}

# * Load world map ----
sf::sf_use_s2(FALSE)
global_land_mask <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  sf::st_as_sfc() %>%
  sf::st_union() %>%
  sf::st_make_valid() %>%
  sf::st_wrap_dateline()

global_ocean_mask <- invert_polygon(global_land_mask)

# * Setup ----
# Input files
pa_files <- c(
  "../../glossa_toy_data/pa_data/Natator_depressus.csv",
  "../../glossa_toy_data/pa_data/Dermochelys_coriacea.csv",
  "../../glossa_toy_data/pa_data/0090963-240229165702484.csv",
  "../../glossa_toy_data/pa_data/0090963-240229165702484.csv"
)

historical_files <- "../../glossa_toy_data/historical_layers.zip"

future_files <- c(
  "../../glossa_toy_data/future1_layers.zip",
  "../../glossa_toy_data/future2_layers.zip"
)

# Verify file paths for input files
input_files <- c(pa_files, historical_files, future_files)

for (file_path in input_files) {
  if (!file.exists(file_path)) {
    stop(paste("Error: File", file_path, "not found. Please verify the file path."))
  }
}

# Analysis options
round_digits <- TRUE
n_round_digits <- 4
scale_layers <- TRUE

native_range <- c("historical", "past", "future")
suitable_habitat <- c("historical", "past", "future")
other_analysis <- c("functional_responses")

# Initialize empty output
presence_absence_list <- list(raw_pa = NULL, clean_pa = NULL, model_pa = NULL)
covariate_list <- list(historical = NULL, past = NULL, future = NULL)
prediction_results <- list(historical = NULL, past = NULL, future = NULL)
other_results <- list(variable_importance = NULL, response_curve = NULL, cross_validation = NULL)
pa_cutoff <- list(native_range = NULL, suitable_habitat = NULL)
habitat_suitability <- list(historical = NULL, past = NULL, future = NULL)
design_matrix <- list()
```


## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
