---
title: "Introduction to the GLOSSA workflow in R with a toy example"
subtitle: "Version 0.1.0"
author: "Jorge Mestre Tomás"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Introduction to the GLOSSA workflow in R with a toy example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# Introduction

*Briefly introduce GLOSSA, BART and the simplicity of the toy example*

*Overview of GLOSSA workflow image*

# Load dependencies

```{r}
# Load libraries
library(glossa)
library(dplyr)
library(ggplot2)

# Load world map
sf::sf_use_s2(FALSE)
global_land_mask <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  sf::st_as_sfc() %>%
  sf::st_union() %>%
  sf::st_make_valid() %>%
  sf::st_wrap_dateline()
```

# Load presence(/absence) data and environmental layers

In order to run GLOSSA, we need the species occurrences (and optionally, the absences) as well as the covariate layers corresponding to the time points when those presences were recorded.

In this toy example, we are going to predict the distribution of two species (`sp1` and `sp2`) based on two covariates (`x1` and `x2`) across two different future scenarios (`future1_layers.zip` and `future2_layers.zip`). Visit [How to format input data](https://github.com/jmestret/glossa) for a description of the input file formats.

The layers are passed to GLOSSA as a zipped directory, containing one subdirectory for each covariate and one raster layer for each time point. The future layers provided should have the same structure but include as many raster files as there are future time points you wish to predict:

```
historical_layers.zip           future1_layers.zip            future2_layers.zip
|__ x1                          |__ x1                        |__ x1
|   |__ x1_1.tif                |   |__ x1_3A.tif             |   |__ x1_3B.tif
|   |__ x1_2.tif                |__ x2                        |__ x2
|__ x2                              |__ x2_3A.tif                 |__ x2_3B.tif
    |__ x2_1.tif
    |__ x2_2.tif
```

```{r}
# Load presence(/absence) data
pa_files <- c("../inst/extdata/sp1.csv", "../inst/extdata/sp2.csv")
presence_absence_list <- list()
presence_absence_list$raw_pa <- lapply(pa_files, glossa:::load_presence_absence_data)


historical_files <- "../inst/extdata/historical_layers.zip"
future_files <- c("../inst/extdata/future1_layers.zip", "../inst/extdata/future2_layers.zip")
```
```{r, echo = FALSE}
rbind(
  read.table(pa_files[1], sep = "\t", dec = ".", header = TRUE),
  read.table(pa_files[2], sep = "\t", dec = ".", header = TRUE)
) %>% 
  ggplot(data = .) +
  geom_sf(data = global_land_mask, color = "#353839", fill = "antiquewhite") +
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = species)) +
  scale_color_manual(values = c("#F6733A", "#4FA3AB"), name = NULL) +
  theme(
    panel.grid.major = element_line(
      color = gray(.5),
      linetype = "dashed",
      linewidth = 0.5
    ),
    panel.background = element_rect(fill = "white"),
    axis.title = element_blank(),
    legend.position = "bottom"
  )
```

# Clean coordinates

First, we will clean up some unwanted coordinates. This is a straightforward preprocessing step, as we assume the user has already provided the coordinates, whether processed or not, according to their preference. In this step, we focus on removing duplicate coordinates based on a specified coordinate resolution (use `n_digits` to select the number of decimal places in the coordinates). We also remove longitudinal duplicates (duplicated points across time) because we will be fitting the model by aggregating all observations across the years, as if we were creating a single aggregated snapshot of the historical data. Finally, we remove points that fall outside the marine regions according to the `rnaturalearth` world polygon.
